---
title: "2023_10_09_final_figures"
author: "Eric Hoffmeyer"
date: "2023-10-09"
output: html_document
---

Customizing figures for final submission

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
suppressPackageStartupMessages({
library(tidyverse)
library(plotly)
library(GeneTonic)
library(ComplexHeatmap)
library(DESeq2)
library(pheatmap)
library(EnhancedVolcano)
library(tibble)
})
```

# Read in data

```{r load genetonic inputs}
load("genetonic_results/il8_GeneTonic_inputs.Rdata")
de_genes <- read.csv(file = "deseq_results/il8_DESeq2_analysis.csv")
```

```{r dir folders}
folder.name <- "Analysis_2023_10_09"

if(!exists(folder.name)){
  dir.create(folder.name)
}
```

2023_10_27 Update

More requested graphs. To avoid confusion with previous requests, I will make a new subfolder:

```{r}
subfolder.name <- "/2023_10_27_update"

if(!exists(subfolder.name)){
  dir.create(paste0(folder.name, subfolder.name))
}
```

2023_11_21 update

More changes. Another new subfolder:

```{r}
subfolder2.name <- "/2023_11_21_update"

if(!exists(subfolder2.name)){
  dir.create(paste0(folder.name, subfolder2.name))
}
```

## Pathway Summary Plot

```{r pathway plot, fig.width=10}
il8_summary <- gs_summary_overview(res_enrich = il8.topgo.res.sf.df.dds, 
                                   n_gs = 15,
                                   )

o <- il8_summary +
  labs(title = "Pathway Enrichment",
       subtitle = "B7H3-CXCR2 CAR vs B7H3 CAR") +
  theme(text = element_text(size = 15, face = "bold"),
        title = element_text(size = 15, face = "bold", hjust = 0.5),
        axis.title.x = element_text(size = 15, face = "bold"),
        axis.title.y = element_text(size = 15, face = "bold"))
o
```


```{r}
ggsave(filename = paste0(folder.name, "/pathway_enrichment_gene_sets.png"),
       plot = o,
       width = 11,
       height = 8,
       units = "in",
       bg = "white")
```

## Innate Immune Response: gene heatmap

```{r IIR heatmap}
iir_heatmap <- gs_heatmap(se = vst(il8.sf.df.dds),
           res_de = il8.res.sf.df.dds,
           res_enrich = il8.topgo.res.sf.df.dds,
           annotation_obj = il8.sf.df.annotation,
           geneset_id = "GO:0045087",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "Innate Immune Response (GO:0045087)")

#remove labels at bottom
# change color of CAR labels- orange for BC2, blue for B7
#Move B7 samples to left side of plot
# what p cutoff value was used
```

I can't quite figure out how to make changes to this gs_heatmap object. It may be easier to use the raw data and make a pHeatmap. 

First, let me extract the genes of interest for the innate immune response pathway from this gs_heatmap object

```{r}
iir_gene_list <- il8_summary$data[[12,4]] |> 
  strsplit(split = ",")

iir_gene_list <- iir_gene_list[[1]]
```

Isolate the counts from just the significant genes in the innate immune response list.

```{r}
iir_matrix <- de_genes |> 
  na.omit() |> 
  filter(padj <= 0.05) |>
  filter(gene_symbol %in% iir_gene_list) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::select(contains("B7"))
```

Preparing the annotation names and colors here. Jessica's other graphpad figures have B7H3 in orange and B7H3-CXCR2 as blue, so I will assign those to keep the color scheme consistent.

```{r}
b7c2_samples <- c("D1_B7C2_Il8", "D2_B7C2_Il8", "D3_B7C2_Il8")

iir_anno <- data.frame(sample_name = colnames(iir_matrix)) |> 
  mutate(CAR = if_else(sample_name %in% b7c2_samples, "BC2", "B7-H3")) |> 
  column_to_rownames("sample_name")

iir_CAR <- c("#FFCC00", "blue")
names(iir_CAR) <- c("B7-H3", "BC2")
iir_anno_colors <- list(CAR = CAR)
```

Make heatmap

```{r}
p <- pheatmap(iir_matrix, 
         scale = "row",
         fontsize = 12,
         fontsize_row = 15,
         show_colnames = FALSE,
         treeheight_col = 0,
         annotation = anno,
         annotation_colors = anno_colors,
         main = "Innate Immune Response (GO:0045087)")
```

Save heatmap

```{r}
ggsave(filename = paste0(folder.name, "/innate_immune_response_heatmap_row15.png"),
       plot = p,
       width = 7,
       height = 9,
       units = "in")
```

Jess wants a volcano plot of just these genes. We'll need a new dataframe that has the padj and log2FC info.

```{r}
iir_volcano <- de_genes |> 
  dplyr::filter(gene_symbol %in% rownames(iir_matrix))
```

```{r}
w <- EnhancedVolcano(toptable = iir_volcano,
                lab = iir_volcano$gene_symbol,
                x = "log2FoldChange",
                y = "padj",
                xlim =  c(-1, 5.5),
                ylim = c(0, 4),
                title = "B7H3-CXCR2 vs B7H3 CAR T cells",
                subtitle = "Innate Immune Response Genes",
                FCcutoff = 0.25,
                pCutoff = 0.05,
                legendPosition = "none",
                labFace = "bold",
                caption = "",
                max.overlaps = Inf,
                drawConnectors = T
                )
w
```

```{r}
ggsave(filename = paste0(folder.name, subfolder.name, "/BC_vs_B7_IL8_iir_volcano_plot_25_cutoff.png"),
       plot = w,
       width = 10,
       height = 9,
       units = "in",
       dpi = 700
       )
```

## Heatmap for top 25 overall DESeq2 genes

Jessica asked for the "top 25" genes. We will do two heatmaps. One by adjusted p value and another by log2FC (and still p < 0.05).
The DESeq gene list is already sorted by adjusted p value, so we can start there.

```{r}
top_25_padj <- de_genes |> 
  na.omit() |> 
  dplyr::slice(1:25) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::select(contains("B7"))
```

```{r}
top_25_log2 <- de_genes |> 
  na.omit() |> 
  filter(padj <= 0.05) |> 
  arrange(desc(abs(c(log2FoldChange)))) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::select(contains("B7")) |> 
  dplyr::slice(1:25)
```

Top 25 DESeq genes by adjusted p value heatmap

```{r}
q <- pheatmap(top_25_padj,
         scale = "row",
         show_colnames = FALSE,
         treeheight_col = 0,
         annotation = anno,
         annotation_colors = anno_colors,
         main = "Top 25 genes by adjusted p value: IL8 stim")
q
# ggsave(filename = paste0(folder.name, "/top_25_padj_heatmap.png"),
#        plot = q,
#        width = 7,
#        height = 9,
#        units = "in")

ggsave(filename = paste0(folder.name, subfolder.name, "/BC_vs_B7_IL8_condition_top25_padj_heatmap.png"),
       plot = q,
       width = 7,
       height = 9,
       units = "in")

```

Top 25 DESeq genes by fold change.

```{r}
r <- pheatmap(top_25_log2,
         scale = "row",
         fontsize = 11,
         fontsize_row = 16,
         show_colnames = T,
         treeheight_col = 0,
         annotation = anno,
         annotation_colors = anno_colors,
         main = "Top 25 genes by log2FC value: IL8 stim")
r

# ggsave(filename = paste0(folder.name, "/top_25_log2FC_heatmap.png"),
#        plot = r,
#        width = 7,
#        height = 9,
#        units = "in")

ggsave(filename = paste0(folder.name, subfolder.name, "/BC_vs_B7_IL8_condition_top25_log2FC_heatmap.png"),
       plot = r,
       width = 7,
       height = 9,
       units = "in")
```

## Volcano plots for DE genes.

```{r}
v <- EnhancedVolcano(toptable = de_genes,
                lab = de_genes$gene_symbol,
                x = "log2FoldChange",
                y = "padj",
                xlim =  c(-3, 3),
                ylim = c(-0.5, 8),
                title = "B7H3-CXCR2 vs B7H3 CAR T cells",
                subtitle = "IL8 stim condition",
                FCcutoff = 0.25,
                pCutoff = 0.01,
                legendPosition = "none",
                #labFace = "bold",
                caption = ""
                )
v
```

```{r}
ggsave(filename = paste0(folder.name, "/BC_vs_B7_IL8_condition_volcano_plot.png"),
       plot = v,
       width = 7,
       height = 6,
       units = "in",
       dpi = 700
       )
```

# 2023_10_27 Update

More requested graphs. To avoid confusion with previous requests, I will make a new subfolder

```{r}
subfolder.name <- "/2023_10_27_update"

if(!exists(subfolder.name)){
  dir.create(paste0(folder.name, subfolder.name))
}
```

The new requests are:

  1. Top 25 genes, by log2FC and padj, for BC IL8 stim vs BC no stim.

  2. Heatmap or volcano plot of top 25 genes by logFC of BC IL8 stim vs B7 IL8 stim


## Heatmap/Volcano plot for top 25 log2FC BC IL8 vs B7 IL8

de_genes current lists the DESeq2 results for BC Il8 vs B7 IL8. Make a subset of the top 25

```{r}
top_25_volcano <- de_genes |> 
  na.omit() |> 
  filter(padj <= 0.05) |> 
  arrange(desc(abs(c(log2FoldChange)))) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::slice(1:25)
```

```{r}
s <- EnhancedVolcano(toptable = top_25_volcano,
                lab = rownames(top_25_volcano),
                x = "log2FoldChange",
                y = "padj",
                xlim =  c(-3, 3),
                ylim = c(-0.5, 8),
                title = "B7H3-CXCR2 vs B7H3 CAR T cells",
                subtitle = "IL8 stim condition",
                FCcutoff = 0.5,
                pCutoff = 0.05,
                legendPosition = "none",
                labFace = "bold",
                caption = "",
                max.overlaps = Inf,
                drawConnectors = T
                ) 
s
```

```{r}
ggsave(filename = paste0(folder.name, subfolder.name, "/BC_vs_B7_IL8_condition_top25_log2FC_volcano_plot.png"),
       plot = s,
       width = 9,
       height = 8,
       units = "in",
       dpi = 700
       )
```

Bigger version

```{r}
ggsave(filename = paste0(folder.name, subfolder.name, "/BC_vs_B7_IL8_condition_top25_log2FC_volcano_plot_BIGGER.png"),
       plot = s,
       width = 9,
       height = 8,
       units = "in",
       dpi = 700
       )
```

I will use the heatmap generated previously. This task finished.

# 2023_11_21 Update

More requests.

Double checking some data in the no stim condition. First we load the data.

```{r load genetonic inputs}
load("genetonic_results/no_GeneTonic_inputs.Rdata")
no_stim_de_genes <- read.csv(file = "deseq_results/il8_DESeq2_analysis.csv")
```

I had to go back an re-run the DESeq on the no stim condition, because it was not including the CXCR2 gene. Lowering the minimum read threshold to 4 included the control samples.

Top 25 genes by log2foldchange for no stim condition:

## Top 25 genes by logFC heatmap for BC no stim vs B7 no stim

Make variables of top 25 genes by padj and log2FC:

```{r}
no_stim_t25_log2 <- no_stim_de_genes |> 
  na.omit() |> 
  filter(padj <= 0.05) |> 
  arrange(desc(abs(c(log2FoldChange)))) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::slice(1:25)

#dplyr::select(contains("B7")) use this later for heatmap

no_stim_t25_padj <- no_stim_de_genes |> 
  na.omit() |> 
  dplyr::slice(1:25) |> 
  column_to_rownames(var = "gene_symbol")
```

### Heatmaps

Make annotation tables first

```{r}
no_b7c2_samples <- c("D1_B7C2_no", "D2_B7C2_no", "D3_B7C2_no")

no_matrix_log2 <- no_stim_t25_log2 |> 
  dplyr::select(contains("B7"))

no_anno <- data.frame(sample_name = colnames(no_matrix_log2)) |> 
  mutate(CAR = if_else(sample_name %in% no_b7c2_samples, "BC2", "B7-H3")) |> 
  column_to_rownames("sample_name")

no_CAR <- c("#FFCC00", "blue")
names(no_CAR) <- c("B7-H3", "BC2")
no_anno_colors <- list(CAR = CAR)
```

Now we can make the heatmap. We will also try to add some other requested changes:
  New title: B7-H3 vs BC2
             no stimulation
  Move z scale lower down
  Move annotation legend in closer (on top of z-scale now)
  Change CAR names to B7-H3 and BC2

```{r, fig.width=9, fig.height=10}
r <- pheatmap(no_matrix_log2,
         scale = "row",
         fontsize = 20,
         fontsize_row = 18,
         show_colnames = FALSE,
         legend_breaks = c(1,0,-1),
         treeheight_col = 0,
         annotation = no_anno,
         annotation_colors = no_anno_colors,
         cellwidth = 50,
         width = 9,
         height = 10,
         main = "B7-H3 vs BC2
no stimluation")
```
Save plot

```{r}
ggsave(filename = paste0(folder.name, subfolder2.name, "/BC_vs_B7_no_stim_heatmap_log2FC_11_21_23.png"),
       plot = r,
       width = 10,
       height = 12,
       units = "in")
```

### Volcano plots
In case she asks for them later

```{r}
# t <- EnhancedVolcano(toptable = no_stim_t25_log2,
#                 lab = rownames(no_stim_t25_log2),
#                 x = "log2FoldChange",
#                 y = "padj",
#                 # xlim =  c(-1, 3),
#                 # ylim = c(-0.5, 8),
#                 title = "B7H3-CXCR2 vs B7H3 CAR T cells",
#                 subtitle = "no stim condition",
#                 FCcutoff = 0.5,
#                 pCutoff = 0.05,
#                 legendPosition = "none",
#                 #labFace = "bold",
#                 caption = "",
#                 max.overlaps = Inf,
#                 #drawConnectors = T
#                 ) 
# t
# 
# u <- EnhancedVolcano(toptable = no_stim_t25_padj,
#                 lab = rownames(no_stim_t25_padj),
#                 x = "log2FoldChange",
#                 y = "padj",
#                 # xlim =  c(-3, 3),
#                 # ylim = c(-0.5, 8),
#                 title = "B7H3-CXCR2 vs B7H3 CAR T cells",
#                 subtitle = "no stim condition",
#                 FCcutoff = 0.5,
#                 pCutoff = 0.05,
#                 legendPosition = "none",
#                 #labFace = "bold",
#                 caption = ""
#                 )
# u
# 
# ggsave(filename = paste0(folder.name, subfolder.name, "/BC_vs_B7_no_stim_condition_volcano_plot_top25_log2FC.png"),
#        plot = t,
#        width = 7,
#        height = 6,
#        units = "in",
#        dpi = 700
#        )
# 
# ggsave(filename = paste0(folder.name, subfolder.name, "/BC_vs_B7_no_stim_condition_volcano_plot_top25_padj.png"),
#        plot = u,
#        width = 7,
#        height = 6,
#        units = "in",
#        dpi = 700
#        )
```

Replot the IL8 stim data in a similar fasion:

Top 25 DESeq genes by fold change.

```{r}
top_25_log2 <- de_genes |> 
  na.omit() |> 
  filter(padj <= 0.05) |> 
  arrange(desc(abs(c(log2FoldChange)))) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::select(contains("B7")) |> 
  dplyr::slice(1:25)
```

Annotation df for this data:

```{r}
b7c2_samples <- c("D1_B7C2_Il8", "D2_B7C2_Il8", "D3_B7C2_Il8")

anno <- data.frame(sample_name = colnames(top_25_log2)) |> 
  mutate(CAR = if_else(sample_name %in% b7c2_samples, "BC2", "B7-H3")) |> 
  column_to_rownames("sample_name")

CAR <- c("#FFCC00", "blue")
names(CAR) <- c("B7-H3", "BC2")
anno_colors <- list(CAR = CAR)
```

Heatmap time

```{r, fig.width=9, fig.height=10}
z <- pheatmap(top_25_log2,
         scale = "row",
         fontsize = 20,
         fontsize_row = 18,
         show_colnames = FALSE,
         legend_breaks = c(1,0,-1),
         treeheight_col = 0,
         annotation = anno,
         annotation_colors = anno_colors,
         cellwidth = 50,
         width = 9,
         height = 10,
         main = "B7-H3 vs BC2
IL8 stimulation")
```

```{r}
ggsave(filename = paste0(folder.name, subfolder2.name, "/BC_vs_B7_IL8_top25_log2FC_11_21_23.png"),
       plot = z,
       width = 10,
       height = 12,
       units = "in")
```

Formatting Innate Immune Response Heatmap



```{r, fig.width=9, fig.height=10}
p <- pheatmap(iir_matrix,
         scale = "row",
         fontsize = 20,
         fontsize_row = 18,
         show_colnames = FALSE,
         legend_breaks = c(1,0,-1),
         treeheight_col = 0,
         annotation = iir_anno,
         annotation_colors = iir_anno_colors,
         cellwidth = 50,
         width = 9,
         height = 10,
         main = "B7-H3 vs BC2
Innate Immune Response 
(GO:0045087)")
```


```{r}
ggsave(filename = paste0(folder.name, subfolder2.name, "/BC_vs_B7_iir_path_11_21_23.png"),
       plot = p,
       width = 10,
       height = 12,
       units = "in")
```





