---
title: "GeneTonic bead stim exploration: B7C2 vs B7 CAR T cells"
#output: html_notebook
---

# Load Libraries

These libraries are required to perform this RNAseq analysis. Make sure you have install them before begin the analysis.

```{r load libraries}
suppressPackageStartupMessages({
library(tidyverse)
library(plotly)
library(ggrepel)
library(GeneTonic)
library(ComplexHeatmap)
library(DESeq2)
})
```

# Load data

Previously created in `2023_08_16_subset_analysis.`

```{r load genetonic inputs}
load("genetonic_results/bead_GeneTonic_inputs.Rdata")
```

# Run

Now we can launch a GeneTonic browser to look at the data.

```{r bead genetonic}
GeneTonic(dds = bead.sf.df.dds,
          res_de = bead.res.sf.df.dds,
          res_enrich = bead.topgo.res.sf.df.dds,
          annotation_obj = bead.sf.df.annotation,
          project_id = "bead stim")
```

# Pathway overview graphs

After exploring the data in GeneTonic, I want to save some of the plots. To do this, I can get example code to create the plot from the GeneTonic browser. I will copy/paste the code in here and change the relevant variable names.

## Pathway heatmap

```{r pathway heatmap}
bead_gss_mat <- gs_scores(se = vst(bead.sf.df.dds),
                         res_de = bead.res.sf.df.dds,
                         res_enrich = bead.topgo.res.sf.df.dds,
                         annotation_obj = bead.sf.df.annotation)

bead_gss_plot <- gs_scoresheat(bead_gss_mat, n_gs = 15)

bead_gss_plot
```

```{r save path heatmap}
ggsave("plots/bead/bead_pathway_heatmap.png",
       plot = bead_gss_plot,
       device = "png",
       width = 8,
       height = 6,
       units = "in")
```

## Pathway by p value (and z score)

```{r pathway plot}
bead_summary <- gs_summary_overview(res_enrich = bead.topgo.res.sf.df.dds, n_gs = 15)

bead_summary
```

```{r save path plot}
ggsave("plots/bead/bead_pathway_analysis.png",
       plot = bead_summary,
       device = "png",
       width = 8,
       height = 6,
       units = "in",
       bg = "white")
```

# Individual pathway heatmaps

Mike requested heatmaps for genes in specific pathways from the previous graphs. Here I will generate them.

## Placeholder

```{r heatmap}
_heatmap <- gs_heatmap(se = vst(bead.sf.df.dds),
           res_de = bead.res.sf.df.dds,
           res_enrich = bead.topgo.res.sf.df.dds,
           annotation_obj = bead.sf.df.annotation,
           geneset_id = "GO:XXXXXXXX",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "")
```

```{r save heatmap}
png(filename = "plots/bead/_pathway_heatmap.png",
    width = 8,
    height = 6,
    units = "in",
    res = 600)



dev.off()
```




# Further Pathway exploration

```{r save all pathway analysis}
write.csv(file = "genetonic_results/bead_GeneTonic_pathways.csv", bead.topgo.res.sf.df.dds,
          row.names = FALSE)
```










```{r}
sessionInfo()
```