---
title: "2023_10_09_final_figures"
author: "Eric Hoffmeyer"
date: "2023-10-09"
output: html_document
---

Customizing figures for final submission

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
suppressPackageStartupMessages({
library(tidyverse)
library(plotly)
library(ggrepel)
library(GeneTonic)
library(ComplexHeatmap)
library(DESeq2)
library(pheatmap)
})
```

# Read in data

```{r load genetonic inputs}
load("genetonic_results/il8_GeneTonic_inputs.Rdata")
de_genes <- read.csv(file = "deseq_results/il8_DESeq2_analysis.csv")
```

```{r dir folders}
folder.name <- "Analysis_2023_10_09"

if(!exists(folder.name)){
  dir.create(folder.name)
}
```

# Pathway Summary Plot

```{r pathway plot}
il8_summary <- gs_summary_overview(res_enrich = il8.topgo.res.sf.df.dds, 
                                   n_gs = 15,
                                   )

o <- il8_summary +
  labs(title = "Pathway Enrichment",
       subtitle = "B7H3-CXCR2 CAR vs B7H3 CAR")
o
```


```{r}
ggsave(filename = paste0(folder.name, "/pathway_enrichment_gene_sets.png"),
       plot = o,
       width = 11,
       height = 8,
       units = "in",
       bg = "white")
```

# Innate Immune Response: gene heatmap

```{r IIR heatmap}
iir_heatmap <- gs_heatmap(se = vst(il8.sf.df.dds),
           res_de = il8.res.sf.df.dds,
           res_enrich = il8.topgo.res.sf.df.dds,
           annotation_obj = il8.sf.df.annotation,
           geneset_id = "GO:0045087",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "Innate Immune Response (GO:0045087)")

#remove labels at bottom
# change color of CAR labels- orange for BC2, blue for B7
#Move B7 samples to left side of plot
# what p cutoff value was used
```

I can't quite figure out how to make changes to this gs_heatmap object. It may be easier to use the raw data and make a pHeatmap. 

First, let me extract the genes of interest for the innate immune response pathway from this gs_heatmap object

```{r}
iir_gene_list <- il8_summary$data[[12,4]] |> 
  strsplit(split = ",")

iir_gene_list <- iir_gene_list[[1]]
```

Isolate the counts from just the significant genes in the innate immune response list.

```{r}
iir_matrix <- de_genes |> 
  na.omit() |> 
  filter(padj <= 0.05) |>
  filter(gene_symbol %in% iir_gene_list) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::select(contains("B7"))
```

Preparing the annotation names and colors here. Jessica's other graphpad figures have B7H3 in orange and B7H3-CXCR2 as blue, so I will assign those to keep the color scheme consistent.

```{r}
b7c2_samples <- c("D1_B7C2_Il8", "D2_B7C2_Il8", "D3_B7C2_Il8")

anno <- data.frame(sample_name = colnames(iir_matrix)) |> 
  mutate(CAR = if_else(sample_name %in% b7c2_samples, "B7H3-CXCR2", "B7H3")) |> 
  column_to_rownames("sample_name")

CAR <- c("#FFCC00", "blue")
names(CAR) <- c("B7H3", "B7H3-CXCR2")
anno_colors <- list(CAR = CAR)
```

Make heatmap

```{r}
p <- pheatmap(iir_matrix, 
         scale = "row",
         show_colnames = FALSE,
         treeheight_col = 0,
         annotation = anno,
         annotation_colors = anno_colors,
         main = "Innate Immune Response (GO:0045087)")
```

Save heatmap

```{r}
ggsave(filename = paste0(folder.name, "/innate_immune_response_heatmap.png"),
       plot = p,
       width = 7,
       height = 9,
       units = "in")
```


# Heatmap for top 25 overall DESeq2 genes

Jessica asked for the "top 25" genes. We will do two heatmaps. One by adjusted p value and another by log2FC (and still p < 0.05).
The DESeq gene list is already sorted by adjusted p value, so we can start there. I'm boosting the gene level to 26 since we artifically increased CXCR2 expression.

```{r}
top_25_padj <- de_genes |> 
  na.omit() |> 
  dplyr::slice(1:26) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::select(contains("B7"))
  
```


```{r}
top_25_log2 <- de_genes |> 
  na.omit() |> 
  filter(padj <= 0.05) |> 
  arrange(desc(log2FoldChange)) |> 
  column_to_rownames(var = "gene_symbol") |> 
  dplyr::select(contains("B7")) |> 
  dplyr::slice(1:26)
```

Top 25 DESeq genes by adjusted p value heatmap

```{r}
q <- pheatmap(top_25_padj,
         scale = "row",
         show_colnames = FALSE,
         treeheight_col = 0,
         annotation = anno,
         annotation_colors = anno_colors,
         main = "Top 26 genes by adjusted p value: CAR+IL8 stim")
ggsave(filename = paste0(folder.name, "/top_25_padj_heatmap.png"),
       plot = q,
       width = 7,
       height = 9,
       units = "in")
```

Top 25 DESeq genes by fold change.

```{r}
r <- pheatmap(top_25_log2,
         scale = "row",
         show_colnames = FALSE,
         treeheight_col = 0,
         annotation = anno,
         annotation_colors = anno_colors,
         main = "Top 26 genes by log2FC value: CAR+IL8 stim")
ggsave(filename = paste0(folder.name, "/top_25_log2FC_heatmap.png"),
       plot = r,
       width = 7,
       height = 9,
       units = "in")
```


















