---
title: "2023-08-16_bulk_RNAseq_subset_analysis"
author: "Eric Hoffmeyer"
date: "2023-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries

These libraries are required to perform this RNAseq analysis. Make sure you have install them before begin the analysis.

```{r}
library(tidyverse)
library(tidybulk)
library(janitor)
library(plotly)
library(ggrepel)
library(DESeq2)
library(GeneTonic)
library(pcaExplorer)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(topGO)
```

# Data Dictionary

```{r}
dir.names = c(
"quant_files/D1_B7C2_Il8_stim_S8_L002",
"quant_files/D1_B7C2_bead_stim_S14_L002",
"quant_files/D1_B7C2_no_stim_S2_L002",
"quant_files/D1_B7_Il8_stim_S7_L002",
"quant_files/D1_B7_bead_stim_S13_L002",
"quant_files/D1_B7_no_stim_S1_L002",
"quant_files/D2_B7C2_Il8_stim_S10_L002",
"quant_files/D2_B7C2_bead_stim_S16_L002",
"quant_files/D2_B7C2_no_stim_S4_L002",
"quant_files/D2_B7_Il8_stim_S9_L002",
"quant_files/D2_B7_bead_stim_S15_L002",
"quant_files/D2_B7_no_stim_S3_L002",
"quant_files/D3_B7C2_Il8_stim_S12_L002",
"quant_files/D3_B7C2_bead_stim_S18_L002",
"quant_files/D3_B7C2_no_stim_S6_L002",
"quant_files/D3_B7_Il8_stim_S11_L002",
"quant_files/D3_B7_bead_stim_S17_L002",
"quant_files/D3_B7_no_stim_S5_L002"
)
#results.dir <- file.path("deseq_results")
#plots.dir <- file.path("plots")
```

Convert the `vector` to `data.frame`.

```{r}
Data.dictionary <- data.frame(Orig.Name = dir.names, Folder.Name = dir.names) %>% 
  mutate(Folder.Name = str_remove(Folder.Name, pattern = "quant_files/"))
```

A `DataFrame`:

```{r}
Data.dictionary
```

## Transform Dictionary

Perform these steps to transform the `data.frame` into desirable format.

```{r}
Data.dictionary <- Data.dictionary %>% 
  ## Create or duplicate the "Orig.Name" column and called it "Dup"
  mutate(Dup = Folder.Name) %>% 
  ## Breakup (separate) the "Dup" column using "_" as its separator
  separate_wider_delim(Dup, names = c("donor", "CAR", "stim", "line1", "line2", "line3"), delim = "_") %>%
  ## Create a new column called "col.name" by modifying "Orig.Name"
  ## by removing "_S(any digit)_L002"
  mutate(col.name = str_remove(Folder.Name, pattern = "_stim_S([:digit:]+)_L002")) %>%
  # Relocate the new "col.name" column right after the first column; "Org.Name"
  # relocate(col.name, .after = Orig.Name) %>% 
  ## Make donor, CAR, and stim columns a factors
  mutate(donor = as.factor(donor)) %>% 
  mutate(CAR = as.factor(CAR)) %>% 
  mutate(stim = as.factor(stim)) %>% 
  # mutate(col.name = str_c(donor, CAR, stim, sep = "_")) %>% 
  ## Remove line1, line2, line3 columns, as we don't need them.
  dplyr::select(-c("line1", "line2", "line3"))
```

```{r}
Data.dictionary
```

# Transcript-Gene-Pair: `tx2gene_df`

We previously created the gtf, lookup table, and data matrix made from the salmon quant file loop. We will simply load it here.

## Load the saved copy

```{r}
load(file = "sf.df.Rdata")
```

## Define subsets

```{r}
subset.samples <- Data.dictionary$col.name

il8.subset.samples <- dplyr::filter(Data.dictionary, stim == "Il8")
il8.subset.samples <- il8.subset.samples$col.name

bead.subset.samples <- dplyr::filter(Data.dictionary, stim == "bead")
bead.subset.samples <- bead.subset.samples$col.name

no.subset.samples <- dplyr::filter(Data.dictionary, stim == "no")
no.subset.samples <- no.subset.samples$col.name

```

## (Re)define meta data

```{r}
Data.dictionary.final <- Data.dictionary %>% 
  dplyr::filter(col.name %in% subset.samples)

il8.data.dictionary <- dplyr::filter(Data.dictionary, col.name %in% il8.subset.samples)

bead.data.dictionary <- dplyr::filter(Data.dictionary, col.name %in% bead.subset.samples)

no.data.dictionary <- dplyr::filter(Data.dictionary, col.name %in% no.subset.samples)
```

# Subset and filter count data

```{r}
il8.sf.df.data <- sf.df %>%
  dplyr::select(all_of(il8.subset.samples)) %>%
  drop_na() %>% 
  ## Filter low expressors (<10)
  mutate(Rowsum = apply(., 1, function(x) sum(x > 10))) %>% 
  dplyr::filter(Rowsum == dim(il8.data.dictionary)[1]) %>% 
  ## No longer need the column "Rowsum", we can remove it
  dplyr::select(-Rowsum)

bead.sf.df.data <- sf.df %>%
  dplyr::select(all_of(bead.subset.samples)) %>%
  drop_na() %>% 
  ## Filter low expressors (<10)
  mutate(Rowsum = apply(., 1, function(x) sum(x > 10))) %>% 
  dplyr::filter(Rowsum == dim(bead.data.dictionary)[1]) %>% 
  ## No longer need the column "Rowsum", we can remove it
  dplyr::select(-Rowsum)

no.sf.df.data <- sf.df %>%
  dplyr::select(all_of(no.subset.samples)) %>%
  drop_na() %>% 
  ## Filter low expressors (<10)
  mutate(Rowsum = apply(., 1, function(x) sum(x > 10))) %>% 
  dplyr::filter(Rowsum == dim(no.data.dictionary)[1]) %>% 
  ## No longer need the column "Rowsum", we can remove it
  dplyr::select(-Rowsum)
```

# DESeq Object

## Create SummarizeExperiment Object

```{r}
il8.sf.df.se <- SummarizedExperiment(assays = list(counts = as.matrix(il8.sf.df.data)),
                                colData = il8.data.dictionary)

bead.sf.df.se <- SummarizedExperiment(assays = list(counts = as.matrix(bead.sf.df.data)),
                                colData = bead.data.dictionary)

no.sf.df.se <- SummarizedExperiment(assays = list(counts = as.matrix(no.sf.df.data)),
                                colData = no.data.dictionary)
```

## Create DESeq Object

```{r}
il8.sf.df.dds <- DESeqDataSet(il8.sf.df.se, design = ~CAR)

bead.sf.df.dds <- DESeqDataSet(bead.sf.df.se, design = ~CAR)

no.sf.df.dds <- DESeqDataSet(no.sf.df.se, design = ~CAR)
```

# PCA Plots

```{r}
il8.sf.df.se.pca <- il8.sf.df.se %>% 
  keep_abundant() %>% 
  scale_abundance() %>% 
  reduce_dimensions(method = "PCA", .dims = 3)
```

## 2D PCA
Set up color palette and theme

```{r}
# Use colourblind-friendly colours
friendly_cols <- dittoSeq::dittoColors()

# Set theme
custom_theme <-
  list(
    scale_fill_manual(values = friendly_cols),
    scale_color_manual(values = friendly_cols),
    theme_bw() +
      theme(
        panel.border = element_blank(),
        axis.line = element_line(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.minor = element_line(size = 0.1),
        text = element_text(size = 9),
        legend.position = "bottom",
        strip.background = element_blank(),
        axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10)),
        axis.title.y = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10)),
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
      )
  )

```

Plot PCA

```{r}
il8.sf.df.se.pca %>% 
  pivot_sample() %>% 
  ggplot(aes(x = PC1, y = PC2, colour = donor)) +
  geom_point() +
  geom_text_repel(aes(label = .sample), show.legend = FALSE, size = 3) +
  custom_theme

```



# DE Analysis

## Statistical Analysis

```{r}
il8.sf.df.dds <- DESeq(il8.sf.df.dds)

bead.sf.df.dds <- DESeq(bead.sf.df.dds)

no.sf.df.dds <- DESeq(no.sf.df.dds)
```

## Analysis Result

```{r}
il8.res.sf.df.dds <- results(il8.sf.df.dds,
                        contrast = c("CAR", "B7C2", "B7"),
                        alpha = 0.05)

bead.res.sf.df.dds <- results(bead.sf.df.dds,
                        contrast = c("CAR", "B7C2", "B7"),
                        alpha = 0.05)

no.res.sf.df.dds <- results(no.sf.df.dds,
                        contrast = c("CAR", "B7C2", "B7"),
                        alpha = 0.05)
```

# Annotation Table

```{r}
il8.sf.df.data.df <- il8.sf.df.data %>% 
  rownames_to_column(var = "gene_id")

bead.sf.df.data.df <- bead.sf.df.data %>% 
  rownames_to_column(var = "gene_id")

no.sf.df.data.df <- no.sf.df.data %>% 
  rownames_to_column(var = "gene_id")
```

```{r}
il8.res.sf.df.dds.annotation <- deseqresult2df(il8.res.sf.df.dds) %>% 
  dplyr::rename(gene_id = id) %>% 
  mutate(gene_symbol = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  mutate(gene_entrez = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first")) %>% 
  mutate(gene_desc = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "GENENAME", multiVals = "first")) %>% 
  relocate(gene_symbol, gene_entrez, gene_desc, .after = gene_id) %>% 
  left_join(il8.sf.df.data.df, by = c("gene_id"="gene_id"))

bead.res.sf.df.dds.annotation <- deseqresult2df(bead.res.sf.df.dds) %>% 
  dplyr::rename(gene_id = id) %>% 
  mutate(gene_symbol = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  mutate(gene_entrez = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first")) %>% 
  mutate(gene_desc = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "GENENAME", multiVals = "first")) %>% 
  relocate(gene_symbol, gene_entrez, gene_desc, .after = gene_id) %>% 
  left_join(bead.sf.df.data.df, by = c("gene_id"="gene_id"))

no.res.sf.df.dds.annotation <- deseqresult2df(no.res.sf.df.dds) %>% 
  dplyr::rename(gene_id = id) %>% 
  mutate(gene_symbol = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  mutate(gene_entrez = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first")) %>% 
  mutate(gene_desc = mapIds(org.Hs.eg.db, keys = gene_id, keytype = "ENSEMBL", column = "GENENAME", multiVals = "first")) %>% 
  relocate(gene_symbol, gene_entrez, gene_desc, .after = gene_id) %>% 
  left_join(no.sf.df.data.df, by = c("gene_id"="gene_id"))
```

```{r}
View(il8.res.sf.df.dds.annotation)
```

## Export Result

Open in Microsoft Excel ...

```{r}
write_csv(il8.res.sf.df.dds.annotation, file = "deseq_results/IL8_DESeq2_analysis.csv")

write_csv(bead.res.sf.df.dds.annotation, file = "deseq_results/Bead_DESeq2_analysis.csv")

write_csv(no.res.sf.df.dds.annotation, file = "deseq_results/No_stim_DESeq2_anaylsis.csv")

```

# Quick Assessemnt

```{r}
sig.gene.cutoff = 0.05 ## Reduce significant gene list only p adjusted significant genes

il8.res.sf.df.dds.annotation %>% 
  dplyr::filter(padj <= sig.gene.cutoff) 

bead.res.sf.df.dds.annotation |>
  dplyr::filter(padj <= sig.gene.cutoff)

no.res.sf.df.dds.annotation |>
  dplyr::filter(padj <= sig.gene.cutoff)
```

# GeneTonic Prep

## GeneTonic Annotation

```{r}
il8.sf.df.annotation <- deseqresult2df(il8.res.sf.df.dds) %>% 
  dplyr::rename(gene_id = id) %>% 
  mutate(gene_name = mapIds(org.Hs.eg.db, 
                            keys = gene_id, 
                            keytype = "ENSEMBL", 
                            column = "SYMBOL", 
                            multiVals = "first")) %>% 
  dplyr::select(gene_id, gene_name) 

bead.sf.df.annotation <- deseqresult2df(bead.res.sf.df.dds) %>% 
  dplyr::rename(gene_id = id) %>% 
  mutate(gene_name = mapIds(org.Hs.eg.db, 
                            keys = gene_id, 
                            keytype = "ENSEMBL", 
                            column = "SYMBOL", 
                            multiVals = "first")) %>% 
  dplyr::select(gene_id, gene_name)

no.sf.df.annotation <- deseqresult2df(no.res.sf.df.dds) %>% 
  dplyr::rename(gene_id = id) %>% 
  mutate(gene_name = mapIds(org.Hs.eg.db, 
                            keys = gene_id, 
                            keytype = "ENSEMBL", 
                            column = "SYMBOL", 
                            multiVals = "first")) %>% 
  dplyr::select(gene_id, gene_name)

```

## Enrichment Analysis

### Extract gene symbol

```{r}
il8.enrich.of.interest <- deseqresult2df(il8.res.sf.df.dds) %>% 
  # rownames_to_column(var = "Ensembl") %>% 
  mutate(Symbol = mapIds(org.Hs.eg.db, keys = id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  dplyr::filter(padj < sig.gene.cutoff) %>% 
  pull(Symbol)

bead.enrich.of.interest <- deseqresult2df(bead.res.sf.df.dds) %>% 
  # rownames_to_column(var = "Ensembl") %>% 
  mutate(Symbol = mapIds(org.Hs.eg.db, keys = id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  dplyr::filter(padj < sig.gene.cutoff) %>% 
  pull(Symbol)

no.enrich.of.interest <- deseqresult2df(no.res.sf.df.dds) %>% 
  # rownames_to_column(var = "Ensembl") %>% 
  mutate(Symbol = mapIds(org.Hs.eg.db, keys = id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  dplyr::filter(padj < sig.gene.cutoff) %>% 
  pull(Symbol)

#length(enrich.of.interest)
#head(enrich.of.interest)
```

## Background gene symbol

```{r}
il8.bg.ids <- deseqresult2df(il8.res.sf.df.dds) %>% 
  mutate(Symbol = mapIds(org.Hs.eg.db, keys = id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  pull(Symbol)
#length(il8.bg.ids)

bead.bg.ids <- deseqresult2df(bead.res.sf.df.dds) %>% 
  mutate(Symbol = mapIds(org.Hs.eg.db, keys = id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  pull(Symbol)

no.bg.ids <- deseqresult2df(no.res.sf.df.dds) %>% 
  mutate(Symbol = mapIds(org.Hs.eg.db, keys = id, keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")) %>% 
  pull(Symbol)
```

## Perform Pathway Analysis

```{r}
#il8 stim
il8.topgo.res.sf.df.dds = pcaExplorer::topGOtable(il8.enrich.of.interest,
                                              il8.bg.ids,
                                              ontology = "BP",
                                              mapping = "org.Hs.eg.db",
                                              geneID = "symbol")

## Reshape for Genetonic
il8.topgo.res.sf.df.dds = shake_topGOtableResult(il8.topgo.res.sf.df.dds)
il8.topgo.res.sf.df.dds = get_aggrscores(il8.topgo.res.sf.df.dds, il8.res.sf.df.dds, il8.sf.df.annotation)

#Repeat for bead stim
bead.topgo.res.sf.df.dds = pcaExplorer::topGOtable(bead.enrich.of.interest,
                                              bead.bg.ids,
                                              ontology = "BP",
                                              mapping = "org.Hs.eg.db",
                                              geneID = "symbol")

## Reshape for Genetonic
bead.topgo.res.sf.df.dds = shake_topGOtableResult(bead.topgo.res.sf.df.dds)
bead.topgo.res.sf.df.dds = get_aggrscores(bead.topgo.res.sf.df.dds, bead.res.sf.df.dds, bead.sf.df.annotation)

#Repeat for no stim
no.topgo.res.sf.df.dds = pcaExplorer::topGOtable(no.enrich.of.interest,
                                              no.bg.ids,
                                              ontology = "BP",
                                              mapping = "org.Hs.eg.db",
                                              geneID = "symbol")

## Reshape for Genetonic
no.topgo.res.sf.df.dds = shake_topGOtableResult(no.topgo.res.sf.df.dds)
no.topgo.res.sf.df.dds = get_aggrscores(no.topgo.res.sf.df.dds, no.res.sf.df.dds, no.sf.df.annotation)
```

```{r}
il8.topgo.res.sf.df.dds

```

# Genetonic

## SAVE Genetonic

```{r}
save(il8.sf.df.dds, il8.res.sf.df.dds, il8.topgo.res.sf.df.dds, il8.sf.df.annotation, 
     file = "genetonic_results/il8_GeneTonic_inputs.Rdata")

save(bead.sf.df.dds, bead.res.sf.df.dds, bead.topgo.res.sf.df.dds, bead.sf.df.annotation, 
     file = "genetonic_results/bead_GeneTonic_inputs.Rdata")

save(no.sf.df.dds, no.res.sf.df.dds, no.topgo.res.sf.df.dds, no.sf.df.annotation, 
     file = "genetonic_results/no_GeneTonic_inputs.Rdata")
```

```{r}
load("GeneTonic_inputs.Rdata")
```

Submit data objects to `Genetonic` web browser

```{r, eval=FALSE}
GeneTonic(dds = il8.sf.df.dds,
          res_de = il8.res.sf.df.dds,
          res_enrich = il8.topgo.res.sf.df.dds,
          annotation_obj = il8.sf.df.annotation,
          project_id = "Il8 stim")

GeneTonic(dds = bead.sf.df.dds,
          res_de = bead.res.sf.df.dds,
          res_enrich = bead.topgo.res.sf.df.dds,
          annotation_obj = bead.sf.df.annotation,
          project_id = "Bead stim")

GeneTonic(dds = no.sf.df.dds,
          res_de = no.res.sf.df.dds,
          res_enrich = no.topgo.res.sf.df.dds,
          annotation_obj = no.sf.df.annotation,
          project_id = "Unstimulated")
```

```{r}
sessionInfo()
```

