---
title: "GeneTonic no stim exploration: B7C2 vs B7 CAR T cells"
#output: html_notebook
---

# Load Libraries

These libraries are required to perform this RNAseq analysis. Make sure you have install them before begin the analysis.

```{r load libraries}
suppressPackageStartupMessages({
library(tidyverse)
library(plotly)
library(ggrepel)
library(GeneTonic)
library(ComplexHeatmap)
library(DESeq2)
library(EnhancedVolcano)
})
```

# Load data

Previously created in `2023_08_16_subset_analysis.`

```{r load genetonic inputs}
load("genetonic_results/no_GeneTonic_inputs.Rdata")
```

# Run

Now we can launch a GeneTonic browser to look at the data.

```{r no genetonic}
GeneTonic(dds = no.sf.df.dds,
          res_de = no.res.sf.df.dds,
          res_enrich = no.topgo.res.sf.df.dds,
          annotation_obj = no.sf.df.annotation,
          project_id = "no stim")
```

# Pathway overview graphs

After exploring the data in GeneTonic, I want to save some of the plots. To do this, I can get example code to create the plot from the GeneTonic browser. I will copy/paste the code in here and change the relevant variable names.

## Pathway heatmap

```{r pathway heatmap}
no_gss_mat <- gs_scores(se = vst(no.sf.df.dds),
                         res_de = no.res.sf.df.dds,
                         res_enrich = no.topgo.res.sf.df.dds,
                         annotation_obj = no.sf.df.annotation)

no_gss_plot <- gs_scoresheat(no_gss_mat, n_gs = 15)

no_gss_plot
```

```{r save path heatmap}
ggsave("plots/no_stim/no_pathway_heatmap.png",
       plot = no_gss_plot,
       device = "png",
       width = 8,
       height = 6,
       units = "in")
```

## Pathway by p value (and z score)

```{r pathway plot}
no_summary <- gs_summary_overview(res_enrich = no.topgo.res.sf.df.dds, n_gs = 15)

no_summary
```

```{r save path plot}
ggsave("plots/no_stim/no_pathway_analysis.png",
       plot = no_summary,
       device = "png",
       width = 8,
       height = 6,
       units = "in",
       bg = "white")
```






# Individual pathway heatmaps


## 

```{r }
XXXX_heatmap <- gs_heatmap(se = vst(no.sf.df.dds),
           res_de = no.res.sf.df.dds,
           res_enrich = no.topgo.res.sf.df.dds,
           annotation_obj = no.sf.df.annotation,
           geneset_id = "GO:XXXXXXXX",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "")
```

```{r }
png(filename = "plots/no_stim/XXXXX_pathway_heatmap.png",
    width = 8,
    height = 6,
    units = "in",
    res = 600)



dev.off()
```




# Further Pathway exploration

```{r save all pathway analysis}
write.csv(file = "genetonic_results/no_stim_GeneTonic_pathways.csv", no.topgo.res.sf.df.dds,
          row.names = FALSE)
```







```{r}
EnhancedVolcano(no.res.sf.df.dds@listData,
                lab = no.sf.df.annotation,
                x = no.res.sf.df.dds@listData$log2FoldChange,
                y = no.res.sf.df.dds@listData$pvalue,
                pCutoff = 0.05,
                FCcutoff = 1.33)

dds <- DESeq(no.sf.df.dds)

no_volcano <- results(dds)
```





























```{r}
sessionInfo()
```