---
title: "GeneTonic Il8 stim exploration: B7C2 vs B7 CAR T cells"
#output: html_notebook
---

# Load Libraries

These libraries are required to perform this RNAseq analysis. Make sure you have install them before begin the analysis.

```{r load libraries}
suppressPackageStartupMessages({
library(tidyverse)
library(plotly)
library(ggrepel)
library(GeneTonic)
library(ComplexHeatmap)
library(DESeq2)
})
```

# Load data

Previously created in `2023_08_16_subset_analysis.`

```{r load genetonic inputs}
load("genetonic_results/il8_GeneTonic_inputs.Rdata")
```

# Run

Now we can launch a GeneTonic browser to look at the data.

```{r il8 genetonic}
GeneTonic(dds = il8.sf.df.dds,
          res_de = il8.res.sf.df.dds,
          res_enrich = il8.topgo.res.sf.df.dds,
          annotation_obj = il8.sf.df.annotation,
          project_id = "Il8 stim")
```

# Pathway overview graphs

After exploring the data in GeneTonic, I want to save some of the plots. To do this, I can get example code to create the plot from the GeneTonic browser. I will copy/paste the code in here and change the relevant variable names.

## Pathway heatmap

```{r pathway heatmap}
il8_gss_mat <- gs_scores(se = vst(il8.sf.df.dds),
                         res_de = il8.res.sf.df.dds,
                         res_enrich = il8.topgo.res.sf.df.dds,
                         annotation_obj = il8.sf.df.annotation)

il8_gss_plot <- gs_scoresheat(il8_gss_mat, n_gs = 15)

il8_gss_plot
```

```{r save path heatmap}
ggsave("plots/il8/il8_pathway_heatmap.png",
       plot = il8_gss_plot,
       device = "png",
       width = 8,
       height = 6,
       units = "in")
```

## Pathway by p value (and z score)

```{r pathway plot}
il8_summary <- gs_summary_overview(res_enrich = il8.topgo.res.sf.df.dds, n_gs = 15)

il8_summary
```

```{r save path plot}
ggsave("plots/il8/il8_pathway_analysis.png",
       plot = il8_summary,
       device = "png",
       width = 8,
       height = 6,
       units = "in",
       bg = "white")
```

# Individual pathway heatmaps

Mike requested heatmaps for genes in specific pathways from the previous graphs. Here I will generate them.

## Positive regulation of IL-12 production

```{r il12 heatmap}
il12_heatmap <- gs_heatmap(se = vst(il8.sf.df.dds),
           res_de = il8.res.sf.df.dds,
           res_enrich = il8.topgo.res.sf.df.dds,
           annotation_obj = il8.sf.df.annotation,
           geneset_id = "GO:0032735",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "Positive Regulation of IL-12 production (GO:0032735)")
```

```{r save il-12 heatmap}
png(filename = "plots/il8/il12_pathway_heatmap.png",
    width = 8,
    height = 6,
    units = "in",
    res = 600)

il12_heatmap

dev.off()
```

## Innate Immune Response Heatmap

```{r IIR heatmap}
iir_heatmap <- gs_heatmap(se = vst(il8.sf.df.dds),
           res_de = il8.res.sf.df.dds,
           res_enrich = il8.topgo.res.sf.df.dds,
           annotation_obj = il8.sf.df.annotation,
           geneset_id = "GO:0045087",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "Innate Immune Response (GO:0045087)")
```

```{r save IIR heatmap}
png(filename = "plots/il8/iir_pathway_heatmap.png",
    width = 8,
    height = 8,
    units = "in",
    res = 600)

iir_heatmap

dev.off()
```

## Cellular Immune Response to Virus

```{r CRTV heatmap}
crtv_heatmap <- gs_heatmap(se = vst(il8.sf.df.dds),
           res_de = il8.res.sf.df.dds,
           res_enrich = il8.topgo.res.sf.df.dds,
           annotation_obj = il8.sf.df.annotation,
           geneset_id = "GO:0098586",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "Cellular Response to Virus (GO:0098586)")

```

```{r save crtv heatmap}
png(filename = "plots/il8/crtv_pathway_heatmap.png",
    width = 8,
    height = 6,
    units = "in",
    res = 600)

crtv_heatmap

dev.off()
```

## Positive Regulation of Myeloid Differentiation

```{r prmld heatmap}
prmld_heatmap <- gs_heatmap(se = vst(il8.sf.df.dds),
           res_de = il8.res.sf.df.dds,
           res_enrich = il8.topgo.res.sf.df.dds,
           annotation_obj = il8.sf.df.annotation,
           geneset_id = "GO:0002763",
           cluster_columns = TRUE,
           anno_col_info = "CAR",
           scale_row = TRUE,
           plot_title = "Positive Regulation of Myeloid Leukocyte Differentiation (GO:0002763)")
```

```{r save prmld heatmap}
png(filename = "plots/il8/prmld_pathway_heatmap.png",
    width = 8,
    height = 6,
    units = "in",
    res = 600)

prmld_heatmap

dev.off()
```


# Further Pathway exploration

```{r save all pathway analysis}
write.csv(file = "genetonic_results/il8_GeneTonic_pathways.csv", il8.topgo.res.sf.df.dds,
          row.names = FALSE)
```


## Telomerase pathways

```{r telo df}
telo.topgo.res.sf.df.dds <- il8.topgo.res.sf.df.dds |>
    dplyr::filter(grepl("telomerase", gs_description))
```

```{r telomerase heatmap}
telo_gss_mat <- gs_scores(se = vst(il8.sf.df.dds),
                         res_de = il8.res.sf.df.dds,
                         res_enrich = telo.topgo.res.sf.df.dds,
                         annotation_obj = il8.sf.df.annotation)

telo_gss_mat <- gs_scoresheat(telo_gss_mat, n_gs = 15)

telo_gss_mat
```

```{r save telo heatmap}
ggsave("plots/il8/telomerase_pathway_heatmap.png",
       plot = telo_gss_mat,
       device = "png",
       width = 8,
       height = 4,
       units = "in")
```

## Metabolism pathways

```{r metab df}
metab.topgo.res.sf.df.dds <- il8.topgo.res.sf.df.dds |>
    dplyr::filter(grepl("metab", gs_description))
```

```{r metabolic heatmap}
metab_gss_mat <- gs_scores(se = vst(il8.sf.df.dds),
                         res_de = il8.res.sf.df.dds,
                         res_enrich = metab.topgo.res.sf.df.dds,
                         annotation_obj = il8.sf.df.annotation)

metab_gss_mat <- gs_scoresheat(metab_gss_mat, n_gs = 15)

metab_gss_mat
```

```{r save metabolic heatmap}
ggsave("plots/il8/metabolic_pathway_heatmap.png",
       plot = metab_gss_mat,
       device = "png",
       width = 8,
       height = 6,
       units = "in")
```

## Immune pathways

```{r immune path df}
immuno.topgo.res.sf.df.dds <- il8.topgo.res.sf.df.dds |>
    dplyr::filter(grepl("immun", gs_description))
```

```{r immuno heatmap}
immuno_gss_mat <- gs_scores(se = vst(il8.sf.df.dds),
                         res_de = il8.res.sf.df.dds,
                         res_enrich = immuno.topgo.res.sf.df.dds,
                         annotation_obj = il8.sf.df.annotation)

immuno_gss_mat <- gs_scoresheat(immuno_gss_mat, n_gs = 15)

immuno_gss_mat
```

```{r save immuno heatmap}
ggsave("plots/il8/immuno_pathway_heatmap.png",
       plot = immuno_gss_mat,
       device = "png",
       width = 8,
       height = 6,
       units = "in")
```


```{r}
rm(il8.deseq.genes)
```


```{r}
EnhancedVolcano(toptable = il8.res.sf.df.dds,
                lab = rownames(il8.res.sf.df.dds),
                x = "log2FoldChange",
                y = "padj",
                xlim =  c(-5, 5),
                ylim = c(-0.5, 25),
                title = "BC2 vs B7 CAR T cells: IL8+CAR stim",
                subtitle = "DESeq2 Results",
                FCcutoff = 0.5
                )
```

```{r}
?EnhancedVolcano
```




























```{r}
sessionInfo()
```